---
title: "More_Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(patchwork)
```

```{r}
df <- read.csv("LengthOfStay_Prod.csv")

df$dialysisrenalendstage <- as.character(df$dialysisrenalendstage)
df$asthma <- as.character(df$asthma)
df$irondef <- as.character(df$irondef)
df$pneum <- as.character(df$pneum)
df$substancedependence <- as.character(df$substancedependence)
df$psychologicaldisordermajor <- as.character(df$psychologicaldisordermajor)
df$depress <- as.character(df$depress)
df$psychother <- as.character(df$psychother)
df$fibrosisandother <- as.character(df$fibrosisandother)
df$malnutrition <- as.character(df$malnutrition)
df$hemo <- as.character(df$hemo)

df$vdate <- as.Date(df$vdate, format="%m/%d/%y")
df$discharged <- as.Date(df$discharged, format="%m/%d/%y")
df$los <- as.numeric(df$discharged - df$vdate)
df$los_char <- as.character(df$los)
```

```{r}
characteristics <- df %>% select(dialysisrenalendstage, asthma, irondef,
                                 pneum, substancedependence, psychologicaldisordermajor,
                                 depress, psychother, fibrosisandother, malnutrition,
                                 hemo)
```

```{r}
patterns <- characteristics %>%
  group_by_all() %>%
  count(name="count", sort=TRUE) %>%
  ungroup %>%
  rownames_to_column("id")

top_patterns <- patterns[1:6,]

top_patterns$percentage <- top_patterns$count / nrow(characteristics)

tidy_patt <- top_patterns %>%
  pivot_longer(!c(id, count, percentage),
               names_to = "characteristic",
               values_to = "yn")
```

```{r, fig.height=6}
p1 <- ggplot(tidy_patt, aes(x=characteristic, y=fct_rev(id), fill=yn)) +
  geom_tile(color="white") +
  scale_fill_manual(values=c("gray81", "plum3")) +
  xlab("Patient Characteristic") +
  ylab("Pattern") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())

p2 <- ggplot(top_patterns, aes(x=percentage, y=fct_rev(id))) +
  geom_col(fill = 'cornflowerblue') +
  xlab('Pattern Count')+
  ylab('') +
  xlim(0, 1) +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

p1 + p2 +
  plot_layout(widths=(c(3,0.8)))
```
- pink squares represent the characteristics that the patient HAS
- the majority of patients did not have any of the characteristics
- of the next 5 most popular patterns (patients do have some characteristics), usually patients only had one of the given characteristics (2, 3, 4, 6), but one pattern (5) had 2 (depression & psychologicaldisordermajor)

```{r}
# need to filter df for the above 6 patterns

# pattern 1: every characteristic == 0
pat1 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[1,2]),
                      asthma==as.integer(top_patterns[1,3]),
                      irondef==as.integer(top_patterns[1,4]),
                      pneum==as.integer(top_patterns[1,5]),
                      substancedependence==as.integer(top_patterns[1,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[1,7]),
                      depress==as.integer(top_patterns[1,8]),
                      psychother==as.integer(top_patterns[1,9]),
                      fibrosisandother==as.integer(top_patterns[1,10]),
                      malnutrition==as.integer(top_patterns[1,11]),
                      hemo==as.integer(top_patterns[1,12])) %>%
  add_column("pattern" = 1)

# pattern 2
pat2 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[2,2]),
                      asthma==as.integer(top_patterns[2,3]),
                      irondef==as.integer(top_patterns[2,4]),
                      pneum==as.integer(top_patterns[2,5]),
                      substancedependence==as.integer(top_patterns[2,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[2,7]),
                      depress==as.integer(top_patterns[2,8]),
                      psychother==as.integer(top_patterns[2,9]),
                      fibrosisandother==as.integer(top_patterns[2,10]),
                      malnutrition==as.integer(top_patterns[2,11]),
                      hemo==as.integer(top_patterns[2,12])) %>%
  add_column("pattern" = 2)

# pattern 3
pat3 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[3,2]),
                      asthma==as.integer(top_patterns[3,3]),
                      irondef==as.integer(top_patterns[3,4]),
                      pneum==as.integer(top_patterns[3,5]),
                      substancedependence==as.integer(top_patterns[3,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[3,7]),
                      depress==as.integer(top_patterns[3,8]),
                      psychother==as.integer(top_patterns[3,9]),
                      fibrosisandother==as.integer(top_patterns[3,10]),
                      malnutrition==as.integer(top_patterns[3,11]),
                      hemo==as.integer(top_patterns[3,12])) %>%
  add_column("pattern" = 3)

# pattern 4
pat4 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[4,2]),
                      asthma==as.integer(top_patterns[4,3]),
                      irondef==as.integer(top_patterns[4,4]),
                      pneum==as.integer(top_patterns[4,5]),
                      substancedependence==as.integer(top_patterns[4,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[4,7]),
                      depress==as.integer(top_patterns[4,8]),
                      psychother==as.integer(top_patterns[4,9]),
                      fibrosisandother==as.integer(top_patterns[4,10]),
                      malnutrition==as.integer(top_patterns[4,11]),
                      hemo==as.integer(top_patterns[4,12])) %>%
  add_column("pattern" = 4)

# pattern 5
pat5 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[5,2]),
                      asthma==as.integer(top_patterns[5,3]),
                      irondef==as.integer(top_patterns[5,4]),
                      pneum==as.integer(top_patterns[5,5]),
                      substancedependence==as.integer(top_patterns[5,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[5,7]),
                      depress==as.integer(top_patterns[5,8]),
                      psychother==as.integer(top_patterns[5,9]),
                      fibrosisandother==as.integer(top_patterns[5,10]),
                      malnutrition==as.integer(top_patterns[5,11]),
                      hemo==as.integer(top_patterns[5,12])) %>%
  add_column("pattern" = 5)

# pattern 6
pat6 <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[6,2]),
                      asthma==as.integer(top_patterns[6,3]),
                      irondef==as.integer(top_patterns[6,4]),
                      pneum==as.integer(top_patterns[6,5]),
                      substancedependence==as.integer(top_patterns[6,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[6,7]),
                      depress==as.integer(top_patterns[6,8]),
                      psychother==as.integer(top_patterns[6,9]),
                      fibrosisandother==as.integer(top_patterns[6,10]),
                      malnutrition==as.integer(top_patterns[6,11]),
                      hemo==as.integer(top_patterns[6,12])) %>%
  add_column("pattern" = 6)

df2 <- rbind(pat1, pat2, pat3, pat4, pat5, pat6)

df2$pattern <- as.character(df2$pattern)
```

we want to know how characteristics/physiological state affect length of patient stay in the hospital

some ideas for plots:
- first look at distribution of lengths of stays, then go from there

```{r}
ggplot(df, aes(x=los)) +
  geom_bar() +
  xlab("Length of Stay (Number of Days)") +
  ylab("Number of Patients") +
  theme_bw()
```
- unimodal, decreasing

now look at the distribution of the lengths of stays for patients with the most common patterns of characteristics
```{r}
ggplot(df2, aes(x=los)) +
  geom_bar() +
  xlab("Length of Stay (Number of Days)") +
  ylab("Number of Patients") +
  theme_bw()
```
- again unimodal, decreasing
- shorter range (1-10 instead of 1-13) - means extreme outliers are not included in the main 6 patterns

compare lengths of stays across different characteristic patterns
```{r}
ggplot(df2, aes(x=los, y=fct_rev(pattern))) +
  geom_boxplot() +
  geom_boxplot(outlier.color = "red", outlier.shape = 1, outlier.size = 3) +
  labs(x="Length of Stay (Number of Days)", y="Characteristics Pattern")
```
- from the above plot, it is clear to see that the length of stay does generally increase if the patient does have some underlying medical conditions
- but it is a uniform increase from a median of ~2 days to ~5 days for characteristics patterns 2-6

compare distribution of LOS across different characteristics patterns
```{r}
ggplot(df2, aes(x = los)) +
  geom_bar() +
  facet_wrap(~pattern)
```
- there are no patients with characteristics patterns 2-6 who stayed for shorter than 3 days

```{r}
df2 %>% filter(pattern != "1") %>%
  ggplot(aes(x = los)) +
  geom_bar() +
  facet_wrap(~pattern)
```

```{r}
ggplot(df, aes(x=hematocrit, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Hematocrit", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=neutrophils, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Neutrophils", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=sodium, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Sodium", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=glucose, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Glucose", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=bloodureanitro, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="BloodUreaNitro", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=creatinine, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="SCreatinine", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=bmi, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="BMI", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=pulse, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Pulse", y="Length of Stay (Number of Days)")
```

```{r}
ggplot(df, aes(x=respiration, y=fct_reorder(los_char, los, .desc=TRUE))) +
  geom_boxplot() +
  labs(x="Respiration", y="Length of Stay (Number of Days)")
```

parallel coordinates plot for seeing physiological data
```{r}
library(GGally)
```

```{r}
pc1 <- df2 %>% filter(pattern=="1") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc2 <- df2 %>% filter(pattern=="2") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc3 <- df2 %>% filter(pattern=="3") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc4 <- df2 %>% filter(pattern=="4") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc5 <- df2 %>% filter(pattern=="5") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc6 <- df2 %>% filter(pattern=="6") %>%
  ggparcoord(columns=16:24, scale="uniminmax", alphaLines=0.1)

pc1 + pc2 + pc3 +
  pc4 + pc5 + pc6 +
  plot_layout(ncol=1)
```






