---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(patchwork)
```

```{r}
df <- read.csv("LengthOfStay.csv")
df <- subset(df, select = -c(eid, discharged, vdate))
df <- df %>% mutate(LengthOfStay = ifelse(lengthofstay>8, "9+", lengthofstay))

df$los_char <- as.character(df$lengthofstay)
```

```{r}
characteristics <- df %>% select(dialysisrenalendstage, asthma, irondef,
                                 pneum, substancedependence, psychologicaldisordermajor,
                                 depress, psychother, fibrosisandother, malnutrition,
                                 hemo)

characteristics$dialysisrenalendstage <- as.character(characteristics$dialysisrenalendstage)
characteristics$asthma <- as.character(characteristics$asthma)
characteristics$irondef <- as.character(characteristics$irondef)
characteristics$pneum <- as.character(characteristics$pneum)
characteristics$substancedependence <- as.character(characteristics$substancedependence)
characteristics$psychologicaldisordermajor <- as.character(characteristics$psychologicaldisordermajor)
characteristics$depress <- as.character(characteristics$depress)
characteristics$psychother <- as.character(characteristics$psychother)
characteristics$fibrosisandother <- as.character(characteristics$fibrosisandother)
characteristics$malnutrition <- as.character(characteristics$malnutrition)
characteristics$hemo <- as.character(characteristics$hemo)
```

```{r}
patterns <- characteristics %>%
  group_by_all() %>%
  count(name="count", sort=TRUE) %>%
  ungroup %>%
  rownames_to_column("id")

top_patterns <- patterns[1:8,]

top_patterns$percentage <- top_patterns$count / nrow(characteristics)

tidy_patt <- top_patterns %>%
  pivot_longer(!c(id, count, percentage),
               names_to = "characteristic",
               values_to = "yn")
```

```{r, fig.height=6}
p1 <- ggplot(tidy_patt, aes(x=characteristic, y=fct_rev(id), fill=yn)) +
  geom_tile(color="white") +
  scale_fill_manual(values=c("gray81", "plum3")) +
  xlab("Patient Characteristic") +
  ylab("Pattern") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())

p2 <- ggplot(top_patterns, aes(x=percentage, y=fct_rev(id))) +
  geom_col(fill = 'cornflowerblue') +
  xlab('Pattern Count')+
  ylab('') +
  xlim(0, 1) +
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

p1 + p2 +
  plot_layout(widths=(c(3,1)))
```
- pink squares represent the characteristics that the patient HAS
- the majority of patients did not have any of the characteristics
- of the next 7 most popular patterns (patients do have some characteristics), usually patients only had one of the given characteristics (2, 3, 5, 6, 7), but two patterns (4, 8) had 2

```{r}
find_rows <- function(df, k) {
  pat <- df %>% filter(dialysisrenalendstage==as.integer(top_patterns[k,2]),
                      asthma==as.integer(top_patterns[k,3]),
                      irondef==as.integer(top_patterns[k,4]),
                      pneum==as.integer(top_patterns[k,5]),
                      substancedependence==as.integer(top_patterns[k,6]),
                      psychologicaldisordermajor==as.integer(top_patterns[k,7]),
                      depress==as.integer(top_patterns[k,8]),
                      psychother==as.integer(top_patterns[k,9]),
                      fibrosisandother==as.integer(top_patterns[k,10]),
                      malnutrition==as.integer(top_patterns[k,11]),
                      hemo==as.integer(top_patterns[k,12])) %>%
    add_column("pattern"=k)
  return(pat)
}
```

```{r}
pat1 <- find_rows(df, 1)
pat2 <- find_rows(df, 2)
pat3 <- find_rows(df, 3)
pat4 <- find_rows(df, 4)
pat5 <- find_rows(df, 5)
pat6 <- find_rows(df, 6)
pat7 <- find_rows(df, 7)
pat8 <- find_rows(df, 8)

df2 <- rbind(pat1, pat2, pat3, pat4, pat5, pat6, pat7, pat8)
df2$pattern <- as.character(df2$pattern)
```

look at the distribution of the lengths of stays for patients with the most common patterns of characteristics
```{r}
ggplot(df2, aes(x=LengthOfStay)) +
  geom_bar() +
  xlab("Length of Stay (Number of Days)") +
  ylab("Number of Patients") +
  theme_bw()
```
unimodal, decreasing

compare lengths of stays across different characteristic patterns
```{r}
ggplot(df2, aes(x=lengthofstay, y=fct_rev(pattern))) +
  geom_boxplot() +
  geom_boxplot(outlier.color = "red", outlier.shape = 1, outlier.size = 3) +
  labs(x="Length of Stay (Number of Days)", y="Characteristics Pattern") +
  theme_bw()
```
from the above plot, it is clear to see that the length of stay does generally increase if the patient does have some underlying medical conditions

- mosaic plot: pattern on x-axis, los on y-axis

```{r}
tidy_char <- df2 %>%
  group_by(LengthOfStay, pattern) %>%
  summarise(Freq=n()) %>%
  ungroup()
```

```{r, fig.height=5}
mosaic(LengthOfStay ~ pattern, df2,
       direction=c("v", "h"))
```


OTHER ANALYSIS
- maybe reduce to just 2 patterns (with/without ANY conditions)
```{r}
# make a function that can reduce to 2 patterns
with_without <- function(df) {
  a <- df %>% rowwise() %>%
    mutate(char = ifelse(sum(c_across(dialysisrenalendstage:hemo)) == 0, "Without", "With"))
  return(a)
}

# use this function to create a new df with patterns 1, 2
two_patt <- with_without(df) %>%
  transform(char_fct=factor(char, levels=c("Without", "With")))
```

```{r}
ggplot(two_patt, aes(x=lengthofstay, y=char)) +
  geom_boxplot() +
  geom_boxplot(outlier.color = "red", outlier.shape = 1, outlier.size = 3) +
  labs(x="Length of Stay (Number of Days)", y="Conditions") +
  theme_bw()
```

```{r}
ggplot(two_patt, aes(x=LengthOfStay)) +
  geom_bar() +
  facet_wrap(~char_fct) +
  xlab("Length of Stay (Number of Days)") +
  ylab("Number of Patients") +
  theme_bw()

# add approximation of density curve ?
```

- plot with number of diseases a patient has

```{r}
df$num_char <- rowSums(df[, c(3:13)])
df <- df %>% mutate(num_cond = ifelse(num_char>5, "6+", num_char))

ggplot(df, aes(x=num_cond)) +
  geom_bar() +
  xlab("Number of Medical Conditions") +
  ylab("Number of Patients") +
  theme_bw()
```
- unimodal, decreases very quickly
- majority of the patients do not have any medical conditions

```{r}
library(vcd)
```

```{r}
mosaic(num_cond ~ LengthOfStay, df,
       direction = c("v, h"))
```

```{r}
count <- df %>%
  group_by(LengthOfStay, num_cond) %>%
  summarise(Freq = n()) %>%
  ungroup()

ggplot(count, aes(x=fct_rev(LengthOfStay), y=Freq, color=num_cond)) +
  geom_point() +
  coord_flip() +
  labs(x="Length of Stay (In Days)", 
       y="Number of Patients", 
       color="Number of Conditions") +
  scale_color_viridis_d(direction=-1) +
  theme_bw()
```

```{r}
ggplot(df, aes(x=LengthOfStay, fill=num_cond)) +
  geom_bar(position="dodge") +
  labs(x="Length of Stay (In Days)",
       y="Number of Patients",
       fill="Number of Conditions") +
  scale_fill_brewer(palette="YlOrRd") +
  theme_bw()
```

- looking at rcount (# of admissions to the hospital)

```{r}
count2 <- df %>%
  group_by(LengthOfStay, rcount) %>%
  summarise(Freq = n()) %>%
  ungroup()

ggplot(count2, aes(x=fct_rev(LengthOfStay), y=Freq, color=rcount)) +
  geom_point() +
  coord_flip() +
  labs(x="Length of Stay (In Days)", 
       y="Number of Patients", 
       color="Number of Previous Admissions") +
  scale_color_viridis_d(direction=-1) +
  theme_bw()
```

```{r}
ggplot(df, aes(x=LengthOfStay, fill=rcount)) +
  geom_bar(position="dodge") +
  labs(x="Length of Stay (In Days",
       y="Number of Patients",
       fill="Number of Previous Admissions") +
  theme_bw()
```

- looking at number of conditions, rcount, and los altogether

```{r, fig.height=3}
count3 <- df %>%
  group_by(LengthOfStay, rcount, num_cond) %>%
  summarise(Freq = n()) %>%
  ungroup()

ggplot(count3, aes(x = fct_rev(LengthOfStay), y = Freq, color = num_cond)) +
  geom_point() +
  coord_flip() +
  facet_wrap(~rcount, ncol = 1) +
  labs(y="Number of Patients",
       x="Length of Stay (In Days)",
       color="Number of Conditions") +
  scale_color_viridis_d(direction=-1) +
  theme_bw(6)
```















